Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-05 summary.tex
introduce alias for Omega_K,  \OmK, think I got all...
re-writing sec 7 in progress

To do so, we generalized previous vertically global linear 
calculations of the VSI \citepalias{nelson13,mcnally14,barker15}, which
assumed isothermal perturbations, to include an energy equation with 
a source term in that relaxes the disk
temperature to its initial value on a timescale
$t_c$. This `thermal relaxation' model allows
the thermodynamic response of the disk to be controlled by varying
$t_c$, and has been employed in previous numerical simulations of the
VSI \citepalias{nelson13}.

@@

For a vertically isothermal, sub-adiabatically stratified disk we
determined the thermal timescale   
\begin{align*}
  t_\mathrm{crit} = \frac{ h |q|}{\gamma-1}\Omega_K^{-1}, 
\end{align*}
below which the VSI operates effectively. While $t_\mathrm{crit}$ was
derived for the lowest order VSI mode --- the `fundamental  
corrugation mode' --- we find from numerical calculations that
$t_\mathrm{crit}$ can be used to characterize the general
stabilization of the VSI by a finite thermal timescale. %That is, all
%VSI growth rates are reduced significantly as $\beta$ is increased
%from 
%zero to $\beta_\mathrm{crit}$.    
We have therefore provided a quantitative measure for the thermal
timescale requirement for the VSI in astrophysical 
disks. The thermal timescale must be shorter than the disk
dynamical timescale by $O( h)$, which is
quite short for thin disks ($ h\ll 1$), because vertical shear is weak.

@@

\subsection{Caveats and outlooks} 
We highlight below some extensions to the present  
linear models in future works.

@@

The convective overstability corresponds to slowly growing epicyclic motions.
It is most effective for $t_c\Omega \sim 1$ and disturbances with
$|k_x|\ll |k_z|$, where $k_z$ is a vertical wavenumber. These
conditions are mutually exclusive with the VSI.  
The maximum growth rate for convective overstability is given by 
\begin{align}
  \sigma_\mathrm{conv,max} = -\frac{N^2_r}{4\Omega}\label{max_conv_gen}
\end{align}
\citep{lyra14}, where $N_r^2 = -(c_p\rho)^{-1}\p_rP\p_rS$ is the
square of the radial buoyancy frequency
(cf. Eq. \ref{nzsq_def}). Thus, growth requires $N_r^2<0$.  
%Evaluating Eq. \ref{max_conv_gen} at the midplane of our disk models,
%we find  
%\begin{align}
%  \sigma_\mathrm{conv,max} =
%  -\frac{ h^2\Omega_K}{4\gamma\Gamma}\left(p+q\right)
%  \left[\left(\gamma-1\right)p-q\right]. \label{max_conv}
%\end{align}
In fact, for the MMSN considered in this work, convective
overstability does not apply because 
$N_r^2>0$.

@@

\subsubsection{Additional physics} 
Depending on where the VSI operates in a
disk, additional effects may become important. For example, in the
outer parts of a PPD, where we estimate the VSI 
operates, disk self-gravity (SG) may play a role. Vertical
SG may favor the VSI by increasing $|\p_z\Omega|$ through
enhancing the density stratification 
(Eq. \ref{vertical_shear}). %sg in pert? no because kx>>1. but maybe for kx~1

>>>>

To do so, we generalized previous vertically global linear 
calculations of the VSI \citepalias{nelson13,mcnally14,barker15}, which
assumed isothermal perturbations, to include an energy equation with 
a source term in that relaxes the disk
temperature to its initial value on a timescale
$t_c$. % This `thermal relaxation' model allows
% the thermodynamic response of the disk to be controlled by varying
% $t_c$, and has been employed in previous numerical simulations of the
% VSI \citepalias{nelson13}.

@@

For a vertically isothermal, sub-adiabatically stratified disk we
determined the thermal timescale   
\begin{align*}
  t_\mathrm{crit} = \frac{ h |q|}{\gamma-1}\OmK^{-1}, 
\end{align*}
below which the VSI operates effectively. While $t_\mathrm{crit}$ was
derived for the lowest order VSI mode --- the `fundamental  
corrugation mode' --- we find from numerical calculations that
$t_\mathrm{crit}$ can be used to characterize the general
stabilization of the VSI by a finite thermal timescale. %That is, all
%VSI growth rates are reduced significantly as $\beta$ is increased
%from 
%zero to $\beta_\mathrm{crit}$.    
We have therefore provided a quantitative measure for the thermal
timescale requirement for the VSI in astrophysical 
disks. The thermal timescale must be shorter than the disk
dynamical timescale by $O( h)$, which is
quite short for thin disks ($ h\ll 1$), because vertical shear is weak.

@@

\subsection{Caveats and outlooks} 
We highlight below some outstanding issues that should be addressed in
future works.

\subsubsection{Viscosity} 
% Depending on where the VSI operates in a
% disk, additional effects may become important. For example, in the
% outer parts of a PPD, where we estimate the VSI 
% operates, disk self-gravity (SG) may play a role. Vertical
% SG may favor the VSI by increasing $|\p_z\Omega|$ through
% enhancing the density stratification 
% (Eq. \ref{vertical_shear}). %sg in pert? no because kx>>1. but maybe for kx~1

However, disturbances with large $\khat$ may be subject to viscous
decay in a real disk. Adopting the standard precription for the
kinematic viscosity $\nu = \alpha c_s H$ \citep{shakura73}, where
$\alpha$ is the dimensionless viscosity parameter, 
the viscous timescale for a perturbation lengthscale 
$l\sim 1/k_x$ is $t_\mathrm{visc} = 1/\alpha\khat^2\Omega $. 
Setting $t_\mathrm{visc}$ to be longer than the
minimum VSI growth timescale $1/h\Omega$ implies 
$\khat^2\lesssim h/\alpha$. Inserting $h \sim 0.05$,
we find only modes with $\khat\lesssim 20$ are relevant for
$\alpha\sim 10^{-4}$ or $\khat\lesssim 70$ for $\alpha\sim
10^{-5}$. \citepalias[These low viscosity levels are required for the
VSI, see][]{nelson13}.

@@

The convective overstability corresponds to slowly growing epicyclic motions.
It is most effective for $t_c\Omega \sim 1$ and disturbances with
$|k_x|\ll |k_z|$, where $k_z$ is a vertical wavenumber. These
conditions are mutually exclusive with the VSI.  
The maximum growth rate for convective overstability is given by 
\begin{align}
  \sigma_\mathrm{conv,max} = -\frac{N^2_r}{4\Omega}\label{max_conv_gen}
\end{align}
\citep{lyra14}, where $N_r^2 = -(c_p\rho)^{-1}\p_rP\p_rS$ is the
square of the radial buoyancy frequency
(cf. Eq. \ref{nzsq_def}). Thus, growth requires $N_r^2<0$.  
%Evaluating Eq. \ref{max_conv_gen} at the midplane of our disk models,
%we find  
%\begin{align}
%  \sigma_\mathrm{conv,max} =
%  -\frac{ h^2\OmK}{4\gamma\Gamma}\left(p+q\right)
%  \left[\left(\gamma-1\right)p-q\right]. \label{max_conv}
%\end{align}
In fact, for the MMSN considered in this work, convective
overstability does not apply because 
$N_r^2>0$.

@@
